Adaptive evolution with aneuploidy and mutation

### 2022-01-10 ms/ms.tex
merged and updated 2n*fixation plots

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 59\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}) and \anwt\ fixed in roughly 300 generations on average (green lines in \autoref{fig:ppc-plot}), which agrees with experimental results.
% TODO replace "roughly 300" with the exact number.
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\autoref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing the parameter values from the MAP estimate reduces the model fit to the experimental results~(\autoref{fig:sensitivity}).

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figures~\ref{fig:2n*-fixation} and \ref{fig:likelihood}). The estimated parameter values without aneuploidy are: %% generated by MAP-and-hdi.ipynb
mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
fitness of mutant, $w_{2n^*}=1.013\substack{+0.000 \\ -0.000}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\autoref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO WAIC?

% Fig single-locus model with and without aneuploidy
\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
      \caption{
        \textbf{A. With aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
      \label{fig:likelihood-with-aneuploidy}
  \end{subfigure}

  \begin{subfigure}[t]{0.45\textwidth}
  	  \caption{
        \textbf{B. Without aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-noaneuploidy-plot.pdf}      
      \label{fig:likelihood-without-aneuploidy}
  \end{subfigure}

  \caption{
  % TODO change x label - "Time to fixation of 2n*"
  % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
  % place 82% a bit higher
  % xticks - replace 500, 1000,1500,2000,2500 with 450, 1700, 2350
    \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy.
	}
  \label{fig:likelihood}
\end{figure}

% TODO merge previous and next figure to one figure with 2 panels

%%% generated by with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[h!]
  \centering
\includegraphics[width=0.9\textwidth]{../figures/fixation-plot.pdf}
  \caption{
  \textbf{Adaptation time cannot be explained without aneuploidy.}
The distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200. Thus, the blue distribution demonstrates a better fit compared to the green, and the yellow demonstrates a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
}
  \label{fig:2n*-fixation}
\end{figure}

@@

  \begin{subfigure}{1\textwidth}
    \centering
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p1_mr.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p2_tr.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p3_w1.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p4_w2.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p5_w3.pdf}      
     \label{fig:convergence-A}
    \caption{\textbf{A. Posterior densities per iteration}}
    \end{subfigure}
    \\
    \vspace*{4mm}

  \begin{subfigure}{1\textwidth}
  	\includegraphics[width=1\textwidth]{../figures/ess.pdf}
     \label{fig:convergence-B}
     \vspace*{-8mm}
    \caption{\textbf{B. ABC-SMC population properties per iteration}}
  \end{subfigure}

>>>>

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \autoref{fig:ppc-plot}), which agrees with experimental results.
% TODO replace "roughly 300" with the exact number.
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\autoref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing the parameter values from the MAP estimate reduces the model fit to the experimental results~(\autoref{fig:sensitivity}).

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). The estimated parameter values without aneuploidy are: %% generated by MAP-and-hdi.ipynb
mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
fitness of mutant, $w_{2n^*}=1.013\substack{+0.000 \\ -0.000}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\autoref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO WAIC?

% Fig single-locus model with and without aneuploidy
%%% generated by with-aneuploidy_vs_no_aneuploidy.ipynb

\begin{figure}[b!]

  \begin{subfigure}[t]{0.9\textwidth}
      \centering
      \includegraphics[height=1.8in]{../figures/fixation-plot-a.pdf}      
      \label{fig:fit}
    \caption{\textbf{A. Distribution.}}
  \end{subfigure}

  \vspace*{4mm}

  \begin{subfigure}[t]{0.9\textwidth}
      \centering
      \includegraphics[height=1.8in]{../figures/fixation-plot-b.pdf}      
      \label{fig:fit-cumulative}
    \caption{\textbf{B. Cumulative distribution.}}
  \end{subfigure}

  \caption{
    \textbf{Single-locus model fit with and without aneuploidy.}
    \textbf{(A)} The distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200. Thus, the blue distribution demonstrates a better fit compared to the green, and the yellow demonstrates a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
    \textbf{(B)}
	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy.
	}
  \label{fig:2n*-fixation}
\end{figure}

@@

  \begin{subfigure}{1\textwidth}
    \centering
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p1_mr.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p2_tr.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p3_w1.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p4_w2.pdf}      
      \includegraphics[width=0.45\textwidth]{../figures/convergence-p5_w3.pdf}      
     \label{fig:convergence-A}
    \caption{\textbf{A. Posterior densities per iteration.}}
    \end{subfigure}
    \\
    \vspace*{4mm}

  \begin{subfigure}{1\textwidth}
  	\includegraphics[width=1\textwidth]{../figures/ess.pdf}
     \label{fig:convergence-B}
     \vspace*{-8mm}
    \caption{\textbf{B. ABC-SMC population properties per iteration.}}
  \end{subfigure}
